### 2017-11-03 submission/manuscript.Rmd
complete abstract

**Results.** There was a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness for fecal samples (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study and variable region. Lower Shannon diversity and evennes in fecal samples resulted in a significant increase in relative risk for carinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05). Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).

**Conclusions.**

@@

Our carcinoma observations fit the driver-passenger model and support this concept within the framework of the transition from adenoma to carcinoma. In contrast, the adenoma observations do not fit well with this model and suggests that the transition from control to adenoma do not fit this framework. The stepwise decrease in diversity suggests that the adenoma community is not normal but this change is subtle. Although there may be localized changes that do depend on the driver-passenger model, our observations show that they do not involve the CRC associated genera. It is possible to hypothesize that at early stages of the diease, how the host interacts to these subtle changes could be the catlyst that causes adenoma formation. Subsequent transition to carcinoma could then fit into the proposed driver-passenger model framework.

Although there are still questions that need to be answered for the microbiota and carcinoma a clearer framework is beginning to develop as to how this occurs. The role of the microbiota in adenoma is still not clear and part of the reason may be because many studies are not powered effectely to observe the small changes reported here. It is realistic to suspect that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control. Most of the studies analyzed have sufficient power to detect these changes. In contrast, our data suggests that the adenoma effect size is relatively small. None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls and this may well be the range in which differences occur in adenoma. Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out the role of the microbiota in adenoma genesis.

>>>>

**Results.** There was a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness for fecal samples (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study and variable region. Lower Shannon diversity and evennes in fecal samples resulted in a significant increase in relative risk for carinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05). Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). Using the whole community resulted in a higher classification model AUC for both adenoma and carcinoma for fecal and tissue samples. Most studies are adequately powered for large effect size differences which may be more amenable for carcinoma than for adenoma.

**Conclusions.** This data provides support for the idea of the "driver-passenger" model in both adenoma and carinoma. The data is much stronger for carcinoma and this may in part be due to the low power to detect more subtle changes in the majority of studies that have been performed to date.

@@

Our carcinoma observations fit the driver-passenger model and support this concept within the framework of the transition from adenoma to carcinoma. In contrast, with the present data we can only suggest that the adenoma observations might fit well with this model but the changes that occur at this timepoint are small and possibly focal to the adenoma or specific location. The stepwise decrease in diversity suggests that the adenoma community is not normal but this change is subtle. Although there may be localized changes that do depend on the driver-passenger model, supported by the *Porphyromonas* finding in adenoma stool and the relative total abundance of all four CRC associated genera in tissue, there may be another process involved that ultimately exacerbates the condition from a subtle localized change to a global community one. It is possible to hypothesize that at early stages of the diease, how the host interacts to these subtle changes could be the catlyst that ultimately leads to this larger global dysfunctional community.

Although there are still questions that need to be answered for the microbiota and carcinoma a clearer framework is beginning to develop as to how this occurs. The role of the microbiota in adenoma is still not clear and part of the reason may be because many studies are not powered effectively to observe the small changes reported here. It is realistic to suspect that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control. Most of the studies analyzed have sufficient power to detect these changes. In contrast, our data suggests that the adenoma effect size is relatively small. None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls and this may well be the range in which differences occur in adenoma. Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out the role of the microbiota in adenoma genesis.

### 2017-11-03 submission/manuscript.tex
complete abstract

\textbf{Results.} There was a significant decrease from control to
adenoma to carcinoma for both Shannon diversity and evenness for fecal
samples (P-value \textless{} 0.05) after correcting for study and
variable region. Lower Shannon diversity and evennes in fecal samples
resulted in a significant increase in relative risk for carinoma
(P-value \textless{} 0.05) but not adenoma (P-value \textgreater{}
0.05). Previously associated colorectal cancer genera
(\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, or
\emph{Porphyromonas}) followed a similar pattern with a significantly
increased relative risk by their presence for carcinoma (P-value
\textless{} 0.05) but not adenoma (P-value \textgreater{} 0.05) with the
exception of \emph{Porphyromonas} (P-value \textless{} 0.05).

\textbf{Conclusions.}

@@

Only two measures for adenoma were significant when investigating these
CRC associated genera. The first was \emph{Porphyromonas} for feces (RR
range: 1.06 - 2.29, P-value = 0.023) and the second was the total
relative abundance of all four CRC associated genera on tissue (RR
range: 1.21 - 3.89, P-value = 0.00933)

@@

Our carcinoma observations fit the driver-passenger model and support
this concept within the framework of the transition from adenoma to
carcinoma. In contrast, the adenoma observations do not fit well with
this model and suggests that the transition from control to adenoma do
not fit this framework. The stepwise decrease in diversity suggests that
the adenoma community is not normal but this change is subtle. Although
there may be localized changes that do depend on the driver-passenger
model, our observations show that they do not involve the CRC associated
genera. It is possible to hypothesize that at early stages of the
diease, how the host interacts to these subtle changes could be the
catlyst that causes adenoma formation. Subsequent transition to
carcinoma could then fit into the proposed driver-passenger model
framework.

Although there are still questions that need to be answered for the
microbiota and carcinoma a clearer framework is beginning to develop as
to how this occurs. The role of the microbiota in adenoma is still not
clear and part of the reason may be because many studies are not powered
effectely to observe the small changes reported here. It is realistic to
suspect that many changes in carcinoma could easily result in effect
sizes that are 30\% or more between the case and control. Most of the
studies analyzed have sufficient power to detect these changes. In
contrast, our data suggests that the adenoma effect size is relatively
small. None of the studies analyzed were properly powered to detect a
10\% or lower change between case and controls and this may well be the
range in which differences occur in adenoma. Future studies
investigating adenoma and the microbiota need to take these factors into
consideration if we are to work out the role of the microbiota in
adenoma genesis.

>>>>

\textbf{Results.} There was a significant decrease from control to
adenoma to carcinoma for both Shannon diversity and evenness for fecal
samples (P-value \textless{} 0.05) after correcting for study and
variable region. Lower Shannon diversity and evennes in fecal samples
resulted in a significant increase in relative risk for carinoma
(P-value \textless{} 0.05) but not adenoma (P-value \textgreater{}
0.05). Previously associated colorectal cancer genera
(\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, or
\emph{Porphyromonas}) followed a similar pattern with a significantly
increased relative risk by their presence for carcinoma (P-value
\textless{} 0.05) but not adenoma (P-value \textgreater{} 0.05) with the
exception of \emph{Porphyromonas} (P-value \textless{} 0.05). Using the
whole community resulted in a higher classification model AUC for both
adenoma and carcinoma for fecal and tissue samples. Most studies are
adequately powered for large effect size differences which may be more
amenable for carcinoma than for adenoma.

\textbf{Conclusions.} This data provides support for the idea of the
``driver-passenger'' model in both adenoma and carinoma. The data is
much stronger for carcinoma and this may in part be due to the low power
to detect more subtle changes in the majority of studies that have been
performed to date.

@@

Only two measures for adenoma were significant when investigating these
CRC associated genera. The first was \emph{Porphyromonas} for feces (RR
range: 1.06 - 2.29, P-value = 0.023) and the second was the total
relative abundance of all four CRC associated genera on tissue (RR
range: 1.21 - 3.89, P-value = 0.00933).

@@

Our carcinoma observations fit the driver-passenger model and support
this concept within the framework of the transition from adenoma to
carcinoma. In contrast, with the present data we can only suggest that
the adenoma observations might fit well with this model but the changes
that occur at this timepoint are small and possibly focal to the adenoma
or specific location. The stepwise decrease in diversity suggests that
the adenoma community is not normal but this change is subtle. Although
there may be localized changes that do depend on the driver-passenger
model, supported by the \emph{Porphyromonas} finding in adenoma stool
and the relative total abundance of all four CRC associated genera in
tissue, there may be another process involved that ultimately
exacerbates the condition from a subtle localized change to a global
community one. It is possible to hypothesize that at early stages of the
diease, how the host interacts to these subtle changes could be the
catlyst that ultimately leads to this larger global dysfunctional
community.

Although there are still questions that need to be answered for the
microbiota and carcinoma a clearer framework is beginning to develop as
to how this occurs. The role of the microbiota in adenoma is still not
clear and part of the reason may be because many studies are not powered
effectively to observe the small changes reported here. It is realistic
to suspect that many changes in carcinoma could easily result in effect
sizes that are 30\% or more between the case and control. Most of the
studies analyzed have sufficient power to detect these changes. In
contrast, our data suggests that the adenoma effect size is relatively
small. None of the studies analyzed were properly powered to detect a
10\% or lower change between case and controls and this may well be the
range in which differences occur in adenoma. Future studies
investigating adenoma and the microbiota need to take these factors into
consideration if we are to work out the role of the microbiota in
adenoma genesis.
